---
title: "A study on the differences between age demographics in Canada"
author: "Danny Xu"
date: "October 19, 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## A study on the differences between age demographics in Canada

# Danny Xu
# October 19, 2020

## Abstract

This study is a supplement to other studies that look at Canadian's age and look for correlation between it and other aspects of Canadian life. Here we use the GSS 2017 data, which has a large sample size, and many measured variables. This data set should be able to provide some better insight into this topic than smaller datasets. We will use general linear regression without interaction terms to find statistically significant correlated variables with age. Then the statistically significant parameters will be examined to generate conclusions.

## Introduction

The purpose of this analysis is to relate age with other variables to see how younger people differ from older people with regards to where they live, their mental health, and other variables. This relationship is important because it shows what the government needs to focus on to improve target age demographics. This has value in election strategy because if you know that younger voters have problems and you place a solution to it on your platform, then more young voters will vote for you. Similarly the same can be done for olde demographics. The paper includes an in-depth introduction to the GSS 2017 survey data, a general linear regression model with age as a response variable, and discussion about the conclusions of the model.

## Data

The data is taken from Canadian General Social Survey (GSS) from their 2017 survey. The survey's goal was to gather information about families in Canada. More specifically, socioeconomic status, happiness, and general information about the respondents are contained in this dataset. The data was selected because it could help explain what familial factors are important to happiness.

The dataset is fairly large and was obtained using a well thought out design involving stratification. The data comes from all over Canada, which allows us to make inferences about the entire country. The dataset uses income data from tax forms and other documents which makes those entries extremely accurate, compared to surveys that do not have access to such documents.

The survey was voluntary which could make the dataset a bit biased since the people that did not have the time to answer the survey would be missing in the data. This would be an example of possible non-response bias. In addition, income related questions were found using linking, which the respondent could object to. This means that the income related questions weren't filled out for these people, which can skew the data. Finally, families without telephones would be excluded from the survey since the data was collected over the telephone. Overall, there are a lot of possibilities where non-response bias could arise.

## Questionnaire Discussion

Questionnaire is divided into sections using sentences as a header for each section that describes the section's overall theme. Each question of the survey is concise and to the point, using personal pronouns such as "you". Questions within sections flow off each other, each leading to the next predictably. Vocabulary is simple and should be able to be understood by most people who have a grasp of the language. Questions are mostly either multiple choice, or numeric within a range. This means that the resulting data is much easier to work with than short answer or fill in the blank type questions.

Due to the sheer scale of the study, the questionnaire is very long. Most respondents will probably end up doing the questionnaire over multiple sessions, as it is essentially the size of a small project. The length may definitely put off people from responding who would otherwise respond to a short survey. Sometimes, the complexity of the question does not lend itself to be easily asked as a multiple choice type question. In this case, there ends up being lots of text for each choice and lots of choices available. This makes the questionnaire significantly longer.

## Survey Methodology

The target population of the survey is all the people in the provinces of Canada that are at least 15 years old and are non-institutionalized. The frame population is all the people listed in the land-line and cell phone numbers from the Census and various other sources combined with Statistics Canada's dwelling frame. The sampled population are the people who finished the telephone survey.

They went with a cross-sectional design, which means that they collected data at a specific point in time, looking at various different groups. In this case, they used stratified sampling which is a form of probability sampling. They stratified based on the census metropolitan area (CMA), where each of the biggest cities in Canada had their own strata, and smaller CMAs were combined together depending on closeness into their own separate strata. Non-CMA regions were then grouped by province each into their own strata. In the end, they had 27 strata. A two-stage sampling design was used within each strata. First, telephone numbers were grouped up. Then a person would be identified from these groups, and they would be the only person in their household to be surveyed. For each strata, a sample size was found to make sure that estimates had sufficiently low variability. If the minimum sample sizes were met, the remaining respondents of the sample were chosen such that the estimates would be precise at both the national and the stratum level.

There do not appear to any significant trade offs that were needed to conduct the survey. The length of the survey diminishes the response rate of the survey which is a small trade off. This lower response rate is accounted for by giving every respondent a weighting factor to make sure that the group of sampled respondents are similar to the overall target population. Another minor trade off could be the choice to do the survey over telephone over something like mail. This would leave out non-telephone users from the survey.

The weighting factor helps solve a bit of the non-response bias. In addition to this, they modeled and adjusted for non-response using characteristics of non-response respondents such as income taken from the 2016 Census. Unfortunately we don't have access to the weighting factors which is only available on the microdata file.

## Model

We will be using a general linear regression model on the response variable age. The predictor variables will be familial income, self rated mental health, life feelings, household size, sex, region of Canada, population center type, and citizenship status.

Specifically, we will be using the following equation for each observation $i$:

$$Y_i = \beta_0 + \sum_{j = 1}^{22} (\beta_j X_{ij}) + \varepsilon_i$$

Here, $Y_i$ is representing the value of the ith observation's age.

$\beta_0$ is the intercept parameter. Notice that this is invariant with respect to observations, i.e. it does not change.

$X_{ij}$ are the values of the predictor variables for the ith observation. There are 22 of them which is significantly more than the amount of actual predictor variables we have because some of the predictor variables we have are categorical and thus we need to use dummy variable encoding to actually use linear regression on them. The amount of dummy variables for each categorical predictor variable is $n-1$ where $n$ are the amount of possible values the predictor variable can take.

$\beta_j$ are the parameters associated with each of the 22 predictor variables (including dummy variables).

$\varepsilon_i$ is the error term which represents the noise that data has that our model cannot account for.

Because this is a general linear regression model, we will be using the technique of least squares. That is, we will try to minimize the sum of the squared residuals. The residuals are the difference between the model predicted value of the response variable, denoted $\hat{Y_i}$ and the actual value of $Y_i$. Notice that order in which we take the difference doesn't matter since we are squaring the result.

We use the finite population correction with N = 30633177 using data taken from (3). This is an upper bound on the actual target population as it contains institutionalized people. However, most people are non-institutionalized so the actual target population should not be too far from this value.

The software we use to run the model will be R. We will use the survey package for the actual linear regression. 

Interpolation is relatively safe so we can use the model when our values for the predictor variables are within the minimum and maximum values obtained in the survey. Due to the large sample size of the survey, most of the predictor variables attain values between their theoretical minimum and theoretical maximum with the exception of the cutoffs for the upper end of some variables. These include age, household size, and familial income which aggregate all observations with values higher than a certain number into one category.

Our model always converges because of the lack of non-linear terms, so only one iteration is needed to determine parameter estimates.

Due to the sample size of our data, we do not need to worry too much about regularization since overfitting is a problem that occurs when too many parameters are fitted using not enough data. Our frequency for each category is huge relative to the amount of predictor variables.

We use variance inflation factors (VIFs) to determine multicollinearity in our explanatory variables, rather than checking if each explanatory variable is correlated with another. This is necessary as it is an assumption of linear regression.

We will examine a plot of predicted values generated from the model vs. residuals to determine if the error terms are independent as well as checking for equal variance. We will also use a Q-Q plot to check for normality.

## Results

The plots are generated using ggplot2. Dplyr was used to coerce the data to be nice for some of the plots, such as converting a character vector into factors so that a bar graph could be sorted. 

```{r message = FALSE}
library(dplyr)
library(ggplot2)

# Load the cleaned .csv
gss <- read.csv("gss.csv")
```
```{r message = FALSE, warning = FALSE}
# Plot life feelings
ggplot(gss, aes(feelings_life)) +
  geom_histogram(breaks = c(0,1,2,3,4,5,6,7,8,9,10)) +
  labs(title = "Histogram of Life Feelings",
       x = "Overall Feelings about Life (0-10)",
       y = "Frequency",
       caption = "Figure 1") +
  theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0))

# Summary Stats for life feelings
c("Mean" = mean(gss$feelings_life, na.rm = TRUE),
  "Variance" = var(gss$feelings_life, na.rm = TRUE),
  "Minimum" = min(gss$feelings_life, na.rm = TRUE),
  "Maximum" = max(gss$feelings_life, na.rm = TRUE),
  "Median" = median(gss$feelings_life, na.rm = TRUE))
```
Life feelings appears to have a bimodal distribution, according Figure 1. The average rating was about 8 out of 10.

```{r message = FALSE, warning = FALSE}
# Clean the income data so it can be plotted
income <- gss %>%
  transmute(income_family = case_when(income_family == "Less than $25,000" ~ 12500,
                                      income_family == "$25,000 to $49,999" ~ 37500,
                                      income_family == "$50,000 to $74,999" ~ 62500,
                                      income_family == "$75,000 to $99,999" ~ 87500,
                                      income_family == "$100,000 to $ 124,999" ~ 112500,
                                      income_family == "$125,000 and more" ~ 137500,
                                      TRUE ~ NA_real_))
# Plot income
ggplot(income, aes(income_family)) +
  geom_histogram(breaks = c(0,25000,50000,75000,100000,125000,150000)) +
  labs(title = "Histogram of Family Income",
       x = "Income ($)",
       y = "Frequency",
       caption = "Figure 2") +
  theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0))
```
Family income appears to be bimodal, but due to the nature of the data, the upper bin actually accounts for all respondents above $125000 familial income. So family income is actually unimodal.


```{r message = FALSE, warning = FALSE}
# Change to a factor to order the bar graph
gss$self_rated_mental_health <- factor(gss$self_rated_mental_health,
                                       levels = c("Poor",
                                                  "Fair",
                                                  "Good",
                                                  "Very good",
                                                  "Excellent",
                                                  "Don't know",
                                                  NA))

# Plot the graph
ggplot(gss, aes(self_rated_mental_health)) +
  geom_bar() +
  labs(title = "Bar graph of Self Rated Mental Health",
       x = "Mental Health Rating",
       y = "Frequency",
       caption = "Figure 3") +
  theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0))
```
Self Rated Mental Health is unimodal.

```{r message = FALSE, warning = FALSE}
# Plot age
ggplot(gss, aes(age)) +
  geom_histogram(bins = 30) +
  labs(title = "Histogram of Age",
       x = "Age (years)",
       y = "Frequency",
       caption = "Figure 4") +
  theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0))

# Summary stats of age
c("Mean" = mean(gss$age),
  "variance" = var(gss$age),
  "Minimum" = min(gss$age),
  "Maximum" = max(gss$age),
  "Median" = median(gss$age))
```
Age appears to be trimodal, but the last bin is representative of all ages above 80, so age is actually bimodal. The average age of the respondents was around 52 years old. There seems to be decent representation though for all ages though.


```{r message = FALSE, warning = FALSE}
# Plot household size
ggplot(gss, aes(hh_size)) +
  geom_bar() +
  labs(title = "Bar Graph of Household Size",
       x = "Number of people in household",
       y = "Frequency",
       caption = "Figure 5") +
  theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0))

# Summary stats of household size
c("Mean" = mean(gss$hh_size),
  "Variance" = var(gss$hh_size),
  "Minimum" = min(gss$hh_size),
  "Maximum" = max(gss$hh_size),
  "Median" = median(gss$hh_size))
```
Household size is unimodal with the average household being around 2 people.


```{r message = FALSE, warning = FALSE}
# Plot sex
ggplot(gss, aes(sex)) +
  geom_bar() +
  labs(title = "Bar Graph of Sex",
       x = "Sex",
       y = "Frequency",
       caption = "Figure 6") +
  theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0))
  

```
There seems to be more female respondents than male respondents, although it must be further discussed if this is enough to skew the data.

```{r message = FALSE, warning = FALSE}
# Plot region
ggplot(gss, aes(region)) +
  geom_bar() +
  labs(title = "Bar Graph of Respondent Region",
       x = "Region",
       y = "Frequency",
       caption = "Figure 7") +
  theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0))

```
There is good representation of all the regions according to Figure 9.

```{r message = FALSE, warning = FALSE}
# Plot Population Center type
ggplot(gss, aes(pop_center)) +
  geom_bar() +
  labs(title = "Bar Graph of Respondent Populaton Centre Location",
       x = "Population Centre",
       y = "Frequency",
       caption = "Figure 8") +
  theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0))

```
Unsurprisingly most people are from larger urban centres, but rural areas have a decent representation in the survey.

```{r message = FALSE, warning = FALSE}
# Plot Citizenship Status
ggplot(gss, aes(citizenship_status)) +
  geom_bar() +
  labs(title = "Bar Graph of Citizenship Status",
       x = "Citizenship Status",
       y = "Frequency",
       caption = "Figure 9") +
  theme(plot.title = element_text(hjust = 0.5), plot.caption = element_text(hjust = 0))

```
Most respondents are Canadian Citizens by birth.

```{r message = FALSE, warning = FALSE}
# Load the cleaned .csv again because we modified some columns for plotting purposes above
gss <- read.csv("gss.csv")

library(survey)
library(car)

fpc.srs = rep(30633177,20602)

model.design <- svydesign(id = ~1, data = gss, fpc = fpc.srs)

model <- svyglm(age ~ income_family + self_rated_mental_health + feelings_life + hh_size + sex + region + pop_center + citizenship_status, model.design)


summary(model)

vif(model)
```
Linear regression model results for the model described in the Model section. Also contains VIFs of the predictor variables.

```{r include = FALSE}
# Take the dispersion parameter to find the original model's R squared
summary(svyglm(age~1, design = model.design))

```

```{r echo = FALSE}
1 - 234.1488/314.9525
```
The above value is the $R^2$ of the model, calculated using a null-model. 

```{r echo = FALSE, warning = FALSE}
plot(model, which = c(2,1))
```
Figure 10 - 13. Plots that determine that help determine if the model is justified.

We find that the residuals of the model have constant variance, but the residuals tend to be positive when the predicted age is small. This means that the assumption of independence  of errors is violated for small predicted values. The normality assumption looks fine according to Figure 11, apart from the upper right region, where it stops following the line.

Our $R^2$ for the model is fairly low, due to the wide variations in answers in the predictor variables. It is hard to determine the difference between an 8 or 9 rating on mental health, even more so when it is the respondents themselves who give the rating. The model explains about 25% of the variation in the data, which is weak.

The VIFs which measure multicollinearity are small, so the multicollinearity assumption of linear regression holds for the model. Specifically, a VIF of less than 4 is usually considered acceptable, and we are far below 4 for each explanatory variable.

We found that the poorer respondents were older, with families who made 25000 dollars to 75000 dollars having a statistically significant parameter estimate that was non-zero.

The only statistically significant parameter estimate of self rated mental health was for the poor category, which tells us that respondents with poor mental health were usually younger individuals compared to individuals who didn't respond. The rest of the parameter estimates in this category were not statistically significant.

The better the respondent's overall feeling about life was, the older they usually were. On average, each point in feeling about life had an average increase in age of about 1.

The larger the household size, the younger the respondent tended to be. Every unit increase in household size yielded an average decrease in age of about 6.5.

Male respondents were younger than female respondents by about 0.8 years.

Respondents from the prairies or Quebec tended to be about 1 year younger compared to respondents from the Atlantic.

Rural respondents were on average, older than their urban counterparts by 2.3 years.

Respondents who obtained their citizenship by naturalization were almost 5 years older than the ones who obtained their citizenship by birth.

The rest of the parameter estimates are not statistically significant to have good predictive power.

## Discussion

Overall, we found some interesting trends relating various variables and age. We found that younger people rated their mental health poorly, which could be a sign of a need of more mental health programs in Canada. We did not examine all the variables in the dataset, as we cherry picked a few that seemed worthwhile and were generally less correlated with each other to generate a better statistical model. We wanted to ensure that we had a simple model so that it was easy to find solutions for our regression as well as reduce the complexity of the model itself, since having interaction terms for categorical variables can become quite messy. In doing so, we have no interaction terms in our regression, which limits the scope of our study. The variables we could select from the dataset also do not encapsulate all the important age related variables that we would have liked to explore. Age is something that correlates with fitness, mental capabilities, mood, desires, and many other variables, some of which are extremely hard to phrase as a questionnaire question. Coercing such data is also usually a challenge, as many mind related attributes are things that need to be explained in a short answer section. Demarcating and putting these short answers into categories of some sort is always difficult to accomplish. Even doing such a survey is costly and requires a lot more effort than the one carried out by the GSS 2017 team which avoided short answers possibly for this reason and others.

# Weaknesses

The data definitely seems biased based on many of the figures. For example, in Figure 5, there are 2 distinct peaks of the distribution of age, but the actual population distribution taken from (3) is mostly unimodal and does not have 2 severe peaks in the distribution. In addition, in Figure 6, the distribution of males and females is not likely to have been drawn from a large population that is approximately 50% male, 50% female. It is highly skewed towards females. This hurts our model because sex is one of the explanatory variables with a statistically significant p-value. It is possible that that if the model was using population based data that the sex parameter estimate would be not statistically significant. These things seem to suggest that our data may not be representative of the population, and thus our model should be taken with a grain of salt.

# Next Steps

A future study could look at using logistic regression on one of the binary ordinal variables such as if the self rated mental health is poor or not to see how age impacts that. This can be extended to include any binary version of the many ordinal variables present in our model. Ordinal regression could also be used to get even better insights.

Furthermore, one could obtain the weighting values that were generated by the original survey team that tries to make the sample more representative of the actual target population, and then generating our model or a similar model to determine a better answer than the one in this study.

Additional studies could also focus on finding more variables in the dataset that are correlated strongly with age, and if there are, finding a better statistical model that predicts the age of respondents.

## References

1. “General Social Survey - Family (GSS).” Statistics Canada, www23.statcan.gc.ca/imdb/p2SV.pl?Function=getSurvey.

2. Alexander, R., & Caetano, S., gss_cleaning [Source code].

3. "Population estimates on July 1st, by age and sex." Statistics Canada, https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000501

4. Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2020). dplyr: A Grammar of Data Manipulation. R package version 1.0.2. https://CRAN.R-project.org/package=dplyr
  
5. H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York, 2016.

6. T. Lumley (2020) "survey: analysis of complex survey samples". R package version 4.0.

7. T. Lumley (2004) Analysis of complex survey samples. Journal of Statistical Software 9(1): 1-19

8. T. Lumley (2010) Complex Surveys: A Guide to Analysis Using R. John Wiley and Sons.

9. John Fox and Sanford Weisberg (2019). An {R} Companion to Applied Regression, Third Edition. Thousand Oaks CA: Sage. URL: https://socialsciences.mcmaster.ca/jfox/Books/Companion/

## Appendix

1. https://github.com/valuemeta/STA304-PS2